Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-07-28 sports_analytics.Rmd
light edits

### 4th down strategy
The concept of expected points in American football has many applications.  One of the most notable and well-studied is the use of expected points in the evaluation of 4th down strategy, and there is near universal consensus in the literature that NFL teams have been too conservative when making 4th down decisions.  @Romer2006 examines 4th down decisions in the NFL using expected points by focusing only on example from the first quarter of a game (to avoid issues with end of half and end of game decision making).  They conclude that teams don't go for it enough if they are trying to maximize their probability of winning the game.

Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy. @Romer2006 Examines 4th down decisions in the NFL using expected points.  Focuses only on the first quarter.  Teams don't go for it enough is they are trying to maximize their probability of winning the game. Later work in @Yam2019 used win probability, rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams were to conservative in going for it on 4th down.  In addition, they estimate that a better strategy would have been worth about 0.4 wins per season on average, which is substantial in a season with only 16 games.

@@

### Other applications of expected points in American football
@White2002 presents a tiered logistic regression method that can be, in general, applied to regression setting with a polychotomous response.  Using this technique, they present an example where they estimate the value of NFL plays using a simple expected points model with down, to go, and to goal as predictors.  They then use these results to evaluate and rank NFL quarterbacks.

@Alamar2010 uses an expected points framework to evaluate play calling in the NFL.  However, rather than evaluate each play on its own, the framework presented here evaluates the play in the context of the drive.  Based on play-by-play data from 2005 through 2008, they find that teams are underutilizing passing plays in some situations.

@urschel2011 uses expected points to evaluate NFL coaching decisions on kickoffs.  Specifically, they look at surprise on-sides kicks vs regular kickoff and the decision to accept a touchback vs returning kickoff.  Using data from the 2009 NFL season, the conclude, as many have, that coaches in the NFL tend to make conservative decisions.

@@

```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package. At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
library(chess)
# Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
game() |> plot()

>>>>

### 4th down strategy

The concept of expected points in American football has many applications.  One of the most notable and well-studied is the use of expected points in the evaluation of 4th down strategy, and there is near universal consensus in the literature that NFL teams have been too conservative when making 4th down decisions.  
@Romer2006 examines 4th down decisions in the NFL using expected points by focusing only on examples from the first quarter of a game (to avoid issues with end-of-half and end-of-game decision making). 
They conclude that teams don't go for it enough if they are trying to maximize their probability of winning the game.

Numerous papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy. 
In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams are to conservative in going for it on 4th down. 
In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount in a season with only 16 games.

@@

### Other applications of expected points in American football

@White2002 present a tiered logistic regression method that can be, in general, applied to any regression setting with a polychotomous response.  
Using this technique, they estimate the value of NFL plays using a simple expected points model with down, yards to go, and yards to goal as predictors.
They then use these results to evaluate and rank NFL quarterbacks.

@Alamar2010 uses an expected points framework to evaluate play calling in the NFL. 
However, rather than evaluate each play on its own, they evaluate the play in the context of the drive. 
Based on play-by-play data from 2005 through 2008, they find that teams are underutilizing passing plays in some situations.

@urschel2011 use expected points to evaluate NFL coaching decisions on kickoffs. 
Specifically, they look at surprise on-sides kicks vs. regular kickoff and the decision to accept a touchback vs. returning the kickoff. 
Using data from the 2009 NFL season, the conclude, as many have, that coaches in the NFL tend to make conservative decisions.

@@

```{r chess, eval=whoami::username() == 'bbadumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package. At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
library(chess)
# Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
game() |> plot()
